---
title: "Room simulation"
format: html
execute:
  echo: true
  warning: false
  message: false
---

Let's imagine that dorm room assignments will be out soon, and you REALLY want to be paired with your friend. However, there is a catch: every assignment is completely random, so your seniority (or lack of therefore) or preferences are not taken into account at all. The simulation models this exact scenario: every student is randomly placed into one of the forty room slots, two per room, and we check whether student 1 (you) and student 2 (your friend) happen to share the same number. I am also assuming that in this specific college, there are no single rooms (or 3+ people rooms, by that logic!). Just double rooms.

I am running this simulation 1000 times to give us a more realistic sense of the chances of you being paired with your friend.

The function "same_room_once" randomly assigns 40 students to 20 double rooms and checks if you and your friend share a room.
I chose map_lgl() in order to repeat the random assignment many times and estimate the probability empirically. The function returns TRUE if you and your friend end up in the same room, and FALSE otherwise.
```{r}
#| echo: true
#| message: false
#| warning: false

library(tidyverse)

n_rooms <- 20
room_capacity <- 2
n_students <- 40

you <- 1
friend <- 2

same_room_once <- function(x) {
  assignments <- sample(rep(1:n_rooms, each = room_capacity))
  assignments[you] == assignments[friend]
}

set.seed(47) # I chose 47 as my seed because of Pomona College :) 
n_sims <- 10000

outcomes <- map_lgl(1:n_sims, same_room_once)
prob_same_room <- mean(outcomes)
prob_same_room
```
This code creates a tibble named "room_df" that lists our two possible outcomes, and subsequently assigns each corresponding probability(in "prob_same_room" and its complement "1-prob_same_room"). Then, to make the results more easy to interpretate, I use ggplot to graph the results.
```{r}
room_df <- tibble(
  outcome = c("Same room", "Different rooms"),
  prob = c(prob_same_room, 1 - prob_same_room)
)

ggplot(room_df, aes(x = outcome, y = prob * 100)) +
  geom_col() + geom_text(
    aes(label = paste(round(prob * 100, 2), "%")),
    vjust = -0.5,
    size = 6
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title = "Probability You and Your Friend End Up in the Same Room",
    subtitle = "20 double rooms, 40 students, 10000 random assignments",
    x = NULL,
    y = "Probability (%)"
  )
```
 As I previously mentioned, to estimate these probabilities, I wrote a function that randomly assigns 40 students to 20 rooms, each holding exactly two people, and then checks whether you and your friend share the same room number.The graph shows the probability of each of these events occuring.The insights of this code are a little sad: the chances that you and your friend will get paired up in the same double room are a mere ~2.5%. Therefore, you are way more likely to end up with someone other than your friend (~97.5%). The plot clearly illustrates this imbalance, with the “Different rooms” bar towering over the “Same room” one.