---
title: "Traffic stops"
format: html
---

For this project we will be working with data from the [The Stanford Open Policing Project](https://openpolicing.stanford.edu/). The rich data collected by their team contains millions of traffic stops by law enforcement agencies across the United States. From this dataset, I want to see if there are patterns for what days of the week tend to have more vehicle traffic stops by law enforcement agents in the states of California, Oregon and Washington (west coast!),



First, we need to bring the data through SQL. This is important has this data set is so rich that importing it fully could crash our computers! 
```{r}
library(RMariaDB)
library(tidyverse)
```

```{r}
con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```

```{r}
DBI::dbListTables(con_traffic)
```

Now, let's explore the variables within the tables.
```{sql}
#| connection: con_traffic
SELECT * FROM ca_statewide_2023_01_26 LIMIT 8;
```

```{sql}
#| connection: con_traffic
SELECT * FROM or_statewide_2020_04_01 LIMIT 8;
```

```{sql}
#| connection: con_traffic
SELECT * FROM wa_statewide_2020_04_01 LIMIT 8;
```

We should also take a look into the kinds of variables we are working on for each state.
```{sql}
#| connection: con_traffic
DESCRIBE ca_statewide_2023_01_26;
```

```{sql}
#| connection: con_traffic
DESCRIBE or_statewide_2020_04_01
```

```{sql}
#| connection: con_traffic
DESCRIBE wa_statewide_2020_04_01
```

With all of this information, we now need to join these 3 tables so we can do our west coast analysis.
```{r}
ca_or_wa_weekday <- dbGetQuery(
  con_traffic,
  "
  SELECT
      'California' AS state,
      DAYOFWEEK(date) AS weekday,
      COUNT(*) AS total_stops
  FROM ca_statewide_2023_01_26
  WHERE type = 'vehicular'
  GROUP BY state, DAYOFWEEK(date)

  UNION ALL

  SELECT
      'Oregon' AS state,
      DAYOFWEEK(date) AS weekday,
      COUNT(*) AS total_stops
  FROM or_statewide_2020_04_01
  WHERE type = 'vehicular'
  GROUP BY state, DAYOFWEEK(date)

  UNION ALL

  SELECT
      'Washington' AS state,
      DAYOFWEEK(date) AS weekday,
      COUNT(*) AS total_stops
  FROM wa_statewide_2020_04_01
  WHERE type = 'vehicular'
  GROUP BY state, DAYOFWEEK(date)

  ORDER BY weekday, state;
  "
)
```

Now, we can plot this information!
```{r, warning=FALSE, message=FALSE}
#| warning: false
library(ggplot2)

ggplot(
  ca_or_wa_weekday,
  aes(x = weekday, y = total_stops, color = state, group = state)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:7)+
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = "Traffic Stops by Day of Week (CA, OR, WA)",
        subtitle = "Where 1 = Sun, 2 = Mon, 3 = Tue, 4 = Wed, 5 = Thu, 6 = Fri, 7 = Sat",
    x = "Day of Week",
    y = "Total Stops",
    color = "State"
  )
```
As expected (and this is true for all states considered), out of all weekdays, Friday has the highest traffic stops by enforcement agents. 


Now that we have information on the days of the week, let's be even more precise: what are the most common violations that happen during fridays across these states? Unfortunetly, the variable "violations" is only present in the tables for Washignton and California, so we can't use Oregon here.

First, we need to once more join the tables.

```{r}
friday_violations <- dbGetQuery(
  con_traffic,
  "
  SELECT
      state,
      violation,
      COUNT(*) AS total_violations
  FROM (
      SELECT 
          'California' AS state,
          violation,
          date
      FROM ca_statewide_2023_01_26
      WHERE violation IS NOT NULL AND violation <> ''

      UNION ALL

      SELECT 
          'Washington' AS state,
          violation,
          date
      FROM wa_statewide_2020_04_01
      WHERE violation IS NOT NULL AND violation <> ''
  ) AS combined
  WHERE DAYOFWEEK(date) = 6
  GROUP BY state, violation
  ORDER BY state, total_violations DESC;
  "
)
```

```{r,warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

ordered_df <- top5_friday_violations |>
  mutate(
    state = factor(state, levels = c("California", "Washington")),
    violation = reorder(violation, total_violations)
  )
ggplot(
  ordered_df,
  aes(
    x = total_violations,
    y = violation,
    fill = state
  )
) +
  geom_col(position = "dodge") +
  labs(
    title = "Top 5 friday violations by state (CA and WA)",
    x = "Total Violations",
    y = "Violation",
    fill = "State"
  ) 
```
In both states, speed-related violations seem to be the most common, although California dominates it by far (possibly and substantially because of its much greater population). I would suppose part of the reason why however, in both states, speed-related violations are most common are in part because of people going to party at night and other forms of leisure. These factors can lead to more aggressive driving, higher average speeds, and therefore a greater frequency of speed-related stops.


#References

Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” Nature Human Behaviour, 1–10.
 
Stanford Open Policing Project. (n.d.). OpenPolicing database. Retrieved November 25, 2025, from https://openpolicing.stanford.edu/

