---
title: "Traffic stops"
format: html
freeze: true
author: Lourenço Xavier
date: 1 December, 2025

---

For this project we will be working with data from the [The Stanford Open Policing Project](https://openpolicing.stanford.edu/). The rich data collected by their team contains millions of traffic stops by law enforcement agencies across the United States. From this data set, I want to see if there are patterns for what days of the week tend to have more vehicle traffic stops by law enforcement agents in the states of California, Oregon and Washington (west coast!),



First, we need to bring the data through SQL. This is important as this data set is so rich that importing it fully could crash our computers! 
```{r}
#| echo: true
#| message: false
#| warning: false
library(RMariaDB)
library(tidyverse)
```

```{r}
con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```
First, we print a list of all available tables in this data set so we can choose which ones to analyse
```{r}
DBI::dbListTables(con_traffic)
```

Now, let's explore the variables within the tables.
```{sql}
#| connection: con_traffic
SELECT * FROM ca_statewide_2023_01_26 LIMIT 5;
```

```{sql}
#| connection: con_traffic
SELECT * FROM or_statewide_2020_04_01 LIMIT 5;
```

```{sql}
#| connection: con_traffic
SELECT * FROM wa_statewide_2020_04_01 LIMIT 5;
```

We should also take a look into the kinds of variables we are working on for each state.
```{sql}
#| connection: con_traffic
DESCRIBE ca_statewide_2023_01_26;
```

```{sql}
#| connection: con_traffic
DESCRIBE or_statewide_2020_04_01
```

```{sql}
#| connection: con_traffic
DESCRIBE wa_statewide_2020_04_01
```

With all of this information, we now need to union these 3 tables so we can do our west coast analysis.
```{sql}
#| connection: con_traffic
#| output.var: ca_or_wa_weekday

SELECT 'California' AS state,
       DAYOFWEEK(date) AS weekday,
       COUNT(*) AS total_stops
FROM ca_statewide_2023_01_26
WHERE type = 'vehicular'
GROUP BY weekday

UNION ALL
SELECT 'Oregon',
       DAYOFWEEK(date),
       COUNT(*)
FROM or_statewide_2020_04_01
WHERE type = 'vehicular'
GROUP BY DAYOFWEEK(date)

UNION ALL
SELECT 'Washington',
       DAYOFWEEK(date),
       COUNT(*)
FROM wa_statewide_2020_04_01
WHERE type = 'vehicular'
GROUP BY DAYOFWEEK(date)

ORDER BY weekday, state;
```

Now, we can plot this information!
```{r, warning=FALSE, message=FALSE}
library(ggplot2)

ggplot(ca_or_wa_weekday, aes(weekday, total_stops, color = state)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:7,
                     labels = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")) +
  labs(
    title = "Traffic Stops by Day of Week (CA, OR, WA)",
    x = "Day",
    y = "Stops"
  )
```

As expected (and this is true for all states considered), out of all weekdays, Friday has the highest traffic stops by enforcement agents. This might be because people drive more on Fridays since not only do they go to work, but also go home and then go to other places for leisure. It is hard to say that there is a similar trend going on the other way that is true for all states though (meaning, the day of the week with the lowest traffic stops by enforcement agents!) as there is more variability in this scenario.


Now that we have information on the days of the week, let's be even more precise: what are the most common violations that happen during fridays across these states? Unfortunately, the variable "violations" is only present in the tables for Washignton and California, so we can't use Oregon here. Furthermore, combining Washington and California keeps crashing my computer, so let's focus on just California instead!

First, we need to once more union the tables and also find the top 5 violations on fridays.

```{sql}
#| connection: con_traffic
#| output.var: friday_violations

SELECT
    'California' AS state,
    violation,
    COUNT(*) AS total_violations
FROM ca_statewide_2023_01_26
WHERE violation IS NOT NULL
  AND DAYOFWEEK(date) = 6
GROUP BY violation
ORDER BY total_violations DESC
LIMIT 0,5;
```

Now plotting!
```{r}
#| echo: true
#| message: false
library(dplyr)
library(ggplot2)
ordered_df <- friday_violations |>
  mutate(
    violation = reorder(violation, total_violations)
  )

ggplot(
  ordered_df,
  aes(
    x = total_violations,
    y = violation
  )
) +
  geom_col(fill = "red") +
  labs(
    title = "Top 5 Friday Violations in California",
    x = "Total Violations",
    y = "Violation"
  )
```
In California, speed-related violations seem to be the most common. I would suppose part of the reason why however speed-related violations are most common are in part because of people going to party at night and other forms of leisure. These factors can lead to more aggressive driving, higher average speeds, and therefore a greater frequency of speed-related stops. 


To wrap up this project, let's summarize what I did. I used SQL to query traffic data in all west coast states (not if you consider Alaska to be one though!). After doing this I summarized vehicular stops by traffic agents by the day of the week across all three states, and found that fridays are the days with the highest vehicular stops by traffic agents. Then, I focused on the most common violations that occurred in Fridays in  California, and noticed that the violations in California seem to be of a more safety-concern cause than in Washington, where a lot of violations were associated with procedural violations.

```{r}
dbDisconnect(con_traffic, shutdown=TRUE)
```

*References*

Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” Nature Human Behaviour, 1–10.
 
Stanford Open Policing Project. (n.d.). OpenPolicing database. Retrieved November 25, 2025, from https://openpolicing.stanford.edu/

